Behavior Driven Development in Haskell

### 2012-05-24 paper.tex
Add two notes
re-structuring.

One of the best features of Haskell is its strong type system, 
which ensures no type errors in programs if they can be complied.
Thanks to this, programs written in Haskell have less bugs
than other programming languages which do not provide such strong type systems.

Unfortunately, the strong type system of Haskell does not guarantee
no value errors.
Haskellers is, however, apt to rely on the strong type system too much.
They tend to write less test programs than programmers of other
programming languages. TBD: we need evidence to support this.

This also means that typical usage examples cannot be found from test programs.
We agree that type signatures are good abstract document but
there are many functions whose behaviors cannot be guessed from
their names and type signatures.

Why do we dislike to write test programs?
It is probably because writing test programs is not fun.
To encourage Haskellers to write more test programs
to detect value errors,
we need a scheme to make writing test programs more interesting.

@@

\begin{description}
\item[unit test] tests for individual programs or libraries.
\item[example] Typical usage of functions.
\item[property] Invariants which functions have.
\item[behavior] How do functions behave for users. Behavior can be shown by using both example and property.
\item[design] TBD
\item[documentation] Explanation of functions.
\end{description}

@@

We insist that examples and properties, which are useful for users, should be
documented.
So, let us first observe Haddock, the documentation system of Haskell. Haddock defines annotation syntax in Haskell's comment.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Example annotation}

We can specify examples with the annotation syntax {\tt >>>}.
The following is an example of how to use the {\tt length} function
for lists.

@@

\noindent Example annotations are useful for users but 
it is not used as test case from the developer point of view (before we implemented {\tt doctest}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Property annotation}

Haddock also provides code block annotations indicated with ({\tt >}).
For example, the {\tt container} packages has the following code block annotation:

\begin{quote}
\begin{verbatim}
> Data.Map.null (empty) == True
\end{verbatim}
\end{quote}

\noindent Some Haskellers like this kind of equations and dislike to write it with the example annotation as follows:

\begin{quote}
\begin{verbatim}
>>> Data.Map.null (empty)
True
\end{verbatim}
\end{quote}

We can see the equation as a no-parameter property for {\tt null} and {\tt empty} and
can say that there are potentially demands to show properties with Haddock annotation.
To our experience, it is not difficult to find properties for pure functions.
We believe that Haskellers should show properties of pure functions
in documentation.

But there is no annotation syntax to describe properties in Haddock.
Even if a prefix keyword for properties is defined, 
we have no agreement on how to describe parameterized properties: e.g. how to specify types of funcation parameters. And we don't have automatic test mechanism to use properties in documentation.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Internal documentation}

There are some examples and properties which are not useful for users but useful for developers. One example is a corner case registered in a bug track system. TBD: setup-teardown. We don't have a mechanism to integrate design, documentation and tests.

To specify visible behavior, we propose to use {\tt doctest} integrated with QuickCheck. To describe internal behavior, we propose to use {\tt hspec} aslo integrated with QuickCheck.

To execute examples in haddock annotations, we introduced {\tt doctest} from Python community and implemented it in Haskell.

TBD: command and library. 
TBD: talk about exception.

To specify properties in haddock annotations, we propose a new syntax. In this paper, we use the {\tt prop>} prefix. We can write the no-parameter equation as follows:

\begin{quote}
\begin{verbatim}
prop> Data.Map.null (empty) == True
\end{verbatim}
\end{quote}

@@

\noindent To test properties in haddock annotations, we extended doctest to handle
them with QuickCheck.

@@

@@

undefined.

@@

how to handle error cases.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{doctest and QuickCheck}

@@

\section{Improving QuickCheck}

Since no-parameter properties has no parameters, it is enough to test them
just one.
However, QuickCheck 2.3 or earlier tests no-parameter properties 100 times.
More precisely, the evaluation results are memorized and they are
compared with {\tt True} 100 times.
This will be fixed in the next release of QuickCheck.

@@

We would like to thank
Nick Smallbone for having improved QuickCheck so that no-parameter properties are tested just once.
We are grateful to David Waern and Simon Marlow
for introducing the property syntax to Haddock.

>>>>

One of the best features of Haskell is its strong type system, 
which ensures no type errors in programs if they can be complied.
Thanks to this, it is said that programs written in Haskell have less bugs
than other programming languages which do not provide such strong type systems.

\begin{quote}
    \emph{Would we need evidence to support the claim that "Haskell programs have less bugs"?}
\end{quote}

Unfortunately, the strong type system of Haskell does not guarantee
no value errors.
Haskellers is, however, apt to rely on the strong type system too much.
They tend to write less test code than programmers of other
programming languages.

\begin{quote}
    \emph{we need evidence to support this.}
\end{quote}

This also means that typical usage examples cannot be found from test code.
We agree that type signatures are good abstract document but
there are many functions whose behaviors cannot be guessed from
their names and type signatures.

Why do we dislike to write test code?
It is probably because writing test code is not fun.
To encourage Haskellers to write more test code
to detect value errors,
we need a scheme to make writing test code more interesting.

\begin{quote}
    \emph{We should explain what is BDD around here}
\end{quote}

@@

\begin{description}
\item[unit test] tests for individual programs or libraries.
\item[example] Typical usage of functions.
\item[property] Invariants which functions have.
\item[behavior] How do functions behave for users. Behavior can be shown by using both examples and properties.
\item[design] TBD
\item[documentation] Explanation of functions.
\end{description}

@@

We insist that examples and properties, which are useful for users, should be
documented.
So, let us first observe Haddock (version 2.7.2 or earlier), the documentation system of Haskell. 
Haddock defines annotation syntax in Haskell's comment.

Haddock provides code block annotations indicated with ({\tt >}).
For example, the {\tt container} packages has the following code block annotation:

\begin{quote}
\begin{verbatim}
-- | /O(1)/. Is the map empty?
--
-- > Data.Map.null (empty)           == True
-- > Data.Map.null (singleton 1 'a') == False

null :: Map k a -> Bool
null Tip      = True
null (Bin {}) = False
\end{verbatim}
\end{quote}

TBD: code block cannot show results. properties should be distinguished from examples.

TBD: Talk about test-framework. it integrates HUnit and QuickCheck but not 
suitable for design.

To specify visible behavior, we propose to add two new annotation syntax to Haddock and to use {\tt doctest} integrated with QuickCheck. To describe internal behavior, we propose to use {\tt hspec} aslo integrated with QuickCheck.

Useful behavior should be documented in Haddock. 
For this purpose, we propose the followings:

\begin{itemize}
\item Define a new Haddock annotation syntax for examples.
\item Define a new Haddock annotation syntax for properties.
\item Bring {\tt doctest} from the Python community to the Haskell
community and integrate it with QuickCheck.
\end{itemize}

First, we introduce a new Haddock annotation syntax for the example.
In this paper, we use {\tt >>>} which is the same as Python's one.
Here is an example:

@@

\noindent Also, we can show exceptions.

\begin{quote}
\begin{verbatim}
>>> head []
*** Exception: Prelude.head: empty list
\end{verbatim}
\end{quote}

To specify properties in Haddock annotations, we also propose a new syntax. In this paper, we use the {\tt prop>} prefix. We can write the no-parameter equation as follows:

\begin{quote}
\begin{verbatim}
prop> Data.Map.null empty == True
\end{verbatim}
\end{quote}

@@

Examples and properties in Haddock annotations are automatically
tested by the {\tt doctest} command. We also provided a library
to do this job. Its implementation will discussed in Section XXX.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Haddock}

Haddock version 2.8 or later provides the {\tt >>>} syntax. 
We extended Haddock to support the {\tt prop>} syntax by ourselves
for proof of concept.
The maintainers of Haddock agreed to introduce 
the {\tt prop>} syntax and is now supporting it by their own.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{QuickCheck}

Since no-parameter properties has no parameters, 
it is enough to test them just one.
However, QuickCheck 2.3 or earlier tests no-parameter properties 100 times.
More precisely, the evaluation results are memorized and they are
compared with {\tt True} 100 times.
The current maintainer of QuickCheck modified that
no-parameter properties are tested just once in its source repository.

TBD(@sol): how to obtain examples and properties from Haddock annotations.
Old version uses Haddock API but now uses GHC API directly. Why?

TBD(@sol): escape from HUnit. Why?

TBD(@sol): Old version executes one GHCi for each function examples.
But new version executes one GHCI per module.
Drastic speed up. Do We need a graph for execution time?

TBD: how to handle error cases.

TBD: how to integrate QuickCheck.

TBD.

@@

undefined. (The placeholders library provide better keyword such as {\tt notImplemented} and {\tt todo})

@@

@@

@@

We are grateful to David Waern and Simon Marlow
for introducing two new syntax to Haddock.
We would like to thank
Nick Smallbone for having improved QuickCheck so that no-parameter properties are tested just once.
